Use of the package \texttt{fitdistrplus} to specify a distribution from non-censored or censored data

### 2009-12-17 inst/doc/intro2fitdistrplus.Rnw
Adding of examples to fitdistcens.Rd and various minor changes
last changes before release for CRAN I hope (minor corrections in .Rd, a new paragraph in the vignette and correction of plotdist.R not to cut density plot).

\title{Use of the library \texttt{fitdistrplus} to specify a distribution from 
non-censored or censored data}
\author{Marie Laure Delignette-Muller, Régis Pouillot , Jean-Baptiste Denis and Christophe Dutang}

\begin{document}

@@

Here you will find some easy examples of use of the functions of the library \texttt{fitdistrplus}. 
The aim is to show you by examples how to use these functions to help you to specify a parametric distribution 
from data corresponding to a random sample drawn from a theoretical distribution that you want to describe. 
For details, see the documentation of each function, using the R help command (ex.: \texttt{?fitdist}). 
Do not forget to load the library using the function \texttt{library} before testing following examples.
<<>>=
library(fitdistrplus)
@

@@

<<fig=TRUE>>=
x1<-c(6.4,13.3,4.1,1.3,14.1,10.6,
9.9,9.6,15.3,22.1,13.4,13.2,
8.4,6.3,8.9,5.2,10.9,14.4)
plotdist(x1)
@

@@

<<fig=TRUE>>=
descdist(x1)
@

@@

<<fig=TRUE>>=
descdist(x1,boot=1000)
@

@@

Below is the result of a fit of a gamma distribution by maximum likelihood.
<<fig=TRUE>>=
f1g<-fitdist(x1,"gamma")
plot(f1g)
summary(f1g)
@

@@

Below is the fit of a lognormal distribution.
<<fig=TRUE>>=
f1l<-fitdist(x1,"lnorm")
plot(f1l)
summary(f1l)
@

Below is the fit of a normal distribution.
<<fig=TRUE>>=
f1n<-fitdist(x1,"norm")
plot(f1n)
summary(f1n)
@

Below is the fit of a Weibull distribution.
<<fig=TRUE>>=
f1w<-fitdist(x1,"weibull")
plot(f1w)
summary(f1w)
@

@@

For example, below is the definition of the Gumbel distribution (also named extreme value distribution) and 
a first plot of the data set with the Gumbel distribution with arbitrary values for parameters.
<<fig=TRUE>>=
dgumbel<-function(x,a,b) 1/b*exp((a-x)/b)*exp(-exp((a-x)/b))
pgumbel<-function(q,a,b) exp(-exp((a-q)/b))
qgumbel<-function(p,a,b) a-b*log(-log(p))
plotdist(x1,"gumbel",para=list(a=3,b=2))
@

The same data set may be plotted with a Gumbel distribution with modified values for parameters.
<<fig=TRUE>>=
plotdist(x1,"gumbel",para=list(a=10,b=5))
@

And a Gumbel distribution may be fitted to data with these values for initial parameter values.
<<fig=TRUE>>=
fgu<-fitdist(x1,"gumbel",start=list(a=10,b=5))
plot(fgu)
summary(fgu)
@

@@

Below is an example of the use of this function with the previous fit of the gamma distribution. 
<<fig=TRUE>>=
b1g<-bootdist(f1g)
plot(b1g)
summary(b1g)
@

@@

<<fig=TRUE>>=
x2<-rnbinom(n=100,size=2,prob=0.3)
plotdist(x2,discrete=TRUE)
@

@@

<<fig=TRUE>>=
descdist(x2,discrete=T)
@

@@

Below is the result of the fit of a Poisson distribution with the bootstrap simulations.
<<fig=TRUE>>=
f2p<-fitdist(x2,"pois")
plot(f2p)
summary(f2p)
b2p<-bootdist(f2p)
summary(b2p)
@

Below is the result of the fit of a negative binomial distribution with the boostrap simulations.
<<fig=TRUE>>=
f2n<-fitdist(x2,"nbinom")
plot(f2n)
summary(f2n)
b2n<-bootdist(f2n)
summary(b2n)
@

@@

<<fig=TRUE>>=
d1<-data.frame(
left=c(1.73,1.51,0.77,1.96,1.96,-1.4,-1.4,NA,-0.11,0.55,0.41,
    2.56,NA,-0.53,0.63,-1.4,-1.4,-1.4,NA,0.13),
right=c(1.73,1.51,0.77,1.96,1.96,0,-0.7,-1.4,-0.11,0.55,0.41,
    2.56,-1.4,-0.53,0.63,0,-0.7,NA,-1.4,0.13))
plotdistcens(d1)
@

@@

<<fig=TRUE>>=
d2<-data.frame(left=10^(d1$left),right=10^(d1$right))
plotdistcens(d2,leftNA=0)
@

It is also possible to fix \texttt{rightNA} or \texttt{leftNA} to a realistic 
extreme value, even if not exactly known, to obtain a reasonable global ranking of 
observations, as in the example below for the first dataset.
<<fig=TRUE>>=
plotdistcens(d1,rightNA=3)
@

@@

Below is the result of a fit of a Weibull distribution by maximum likelihood and the results of the 
corresponding boostrap simulations.
<<fig=TRUE>>=
f2w<-fitdistcens(d2,"weibull")
summary(f2w)
plot(f2w,leftNA=0)
@

<<fig=TRUE>>=
b2w<-bootdistcens(f2w)
summary(b2w)
plot(b2w)
@

@@

Below is the fit of a lognormal distribution to the same censored data set.
<<fig=TRUE>>=
f2l<-fitdistcens(d2,"lnorm")
summary(f2l)
plot(f2l,leftNA=0)
@

Below is the fit of an exponential distribution.
<<fig=TRUE>>=
f2e<-fitdistcens(d2,"exp")
summary(f2e)
plot(f2e,leftNA=0)
@

@@

>>>>

\title{Use of the package \texttt{fitdistrplus} to specify a distribution from 
non-censored or censored data}
\author{Marie Laure Delignette-Muller, Régis Pouillot , Jean-Baptiste Denis and Christophe Dutang}

\begin{document}
\setkeys{Gin}{width=0.6\textwidth}

@@

<<echo=FALSE,results=hide>>=
options(digits=3)
@

Here you will find some easy examples of use of the functions of the package \texttt{fitdistrplus}. 
The aim is to show you by examples how to use these functions to help you to specify a parametric distribution 
from data corresponding to a random sample drawn from a theoretical distribution that you want to describe. 
For details, see the documentation of each function, using the R help command (ex.: \texttt{?fitdist}). 
Do not forget to load the package using the function \texttt{library} or \texttt{require}
before testing following examples.
<<>>=
library(fitdistrplus)
@

@@

<<plotdistx1,fig=TRUE>>=
x1<-c(6.4,13.3,4.1,1.3,14.1,10.6,
9.9,9.6,15.3,22.1,13.4,13.2,
8.4,6.3,8.9,5.2,10.9,14.4)
plotdist(x1)
@

@@

<<descdistx1,fig=TRUE>>=
descdist(x1)
@

@@

<<descdistx1boot,fig=TRUE>>=
descdist(x1,boot=1000)
@

@@

Below is the result of a fit of a gamma distribution by maximum likelihood.
<<fitdistx1g,fig=TRUE>>=
f1g<-fitdist(x1,"gamma")
plot(f1g)
summary(f1g)
@

@@

Below is the fit of a lognormal distribution.
<<fitdistx1ln,fig=TRUE>>=
f1l<-fitdist(x1,"lnorm")
plot(f1l)
summary(f1l)
@

Below is the fit of a normal distribution.
<<fitdistx1n,fig=TRUE>>=
f1n<-fitdist(x1,"norm")
plot(f1n)
summary(f1n)
@

Below is the fit of a Weibull distribution.
<<fitdistx1w,fig=TRUE>>=
f1w<-fitdist(x1,"weibull")
plot(f1w)
summary(f1w)
@

@@

For example, below is the definition of the Gumbel distribution (also named extreme value distribution) and 
a first plot of the data set with the Gumbel distribution with arbitrary values for parameters.
<<plotdistx1gu1,fig=TRUE>>=
dgumbel<-function(x,a,b) 1/b*exp((a-x)/b)*exp(-exp((a-x)/b))
pgumbel<-function(q,a,b) exp(-exp((a-q)/b))
qgumbel<-function(p,a,b) a-b*log(-log(p))
plotdist(x1,"gumbel",para=list(a=3,b=2))
@

The same data set may be plotted with a Gumbel distribution with modified values for parameters.
<<plotdistx1gu2,fig=TRUE>>=
plotdist(x1,"gumbel",para=list(a=10,b=5))
@

And a Gumbel distribution may be fitted to data with these values for initial parameter values.
<<fitdistx1gu,fig=TRUE>>=
fgu<-fitdist(x1,"gumbel",start=list(a=10,b=5))
plot(fgu)
summary(fgu)
@

@@

Below is an example of the use of this function with the previous fit of the gamma distribution. 
<<bootdistf1g,fig=TRUE>>=
b1g<-bootdist(f1g)
plot(b1g)
summary(b1g)
@

@@

<<plotdistx2,fig=TRUE>>=
x2<-rnbinom(n=100,size=2,prob=0.3)
plotdist(x2,discrete=TRUE)
@

@@

<<descdistx2,fig=TRUE>>=
descdist(x2,discrete=T)
@

@@

Below is the result of the fit of a Poisson distribution with the bootstrap simulations.
<<fitdistx2p,fig=TRUE>>=
f2p<-fitdist(x2,"pois")
plot(f2p)
summary(f2p)
b2p<-bootdist(f2p)
summary(b2p)
@

Below is the result of the fit of a negative binomial distribution with the boostrap simulations.
<<fitdistx2nb,fig=TRUE>>=
f2n<-fitdist(x2,"nbinom")
plot(f2n)
summary(f2n)
b2n<-bootdist(f2n)
summary(b2n)
@

@@

<<plotdistcensd1,fig=TRUE>>=
d1<-data.frame(
left=c(1.73,1.51,0.77,1.96,1.96,-1.4,-1.4,NA,-0.11,0.55,0.41,
    2.56,NA,-0.53,0.63,-1.4,-1.4,-1.4,NA,0.13),
right=c(1.73,1.51,0.77,1.96,1.96,0,-0.7,-1.4,-0.11,0.55,0.41,
    2.56,-1.4,-0.53,0.63,0,-0.7,NA,-1.4,0.13))
plotdistcens(d1)
@

@@

<<plotdistcensd2,fig=TRUE>>=
d2<-data.frame(left=10^(d1$left),right=10^(d1$right))
plotdistcens(d2,leftNA=0)
@

It is also possible to fix \texttt{rightNA} or \texttt{leftNA} to a realistic 
extreme value, even if not exactly known, to obtain a reasonable global ranking of 
observations, as in the example below for the first dataset.
<<plotdistcensd1bis,fig=TRUE>>=
plotdistcens(d1,rightNA=3)
@

@@

Below is the result of a fit of a Weibull distribution by maximum likelihood and the results of the 
corresponding boostrap simulations.
<<fitdistcensd2w,fig=TRUE>>=
f2w<-fitdistcens(d2,"weibull")
summary(f2w)
plot(f2w,leftNA=0)
@

<<bootdistcensf2w,fig=TRUE>>=
b2w<-bootdistcens(f2w)
summary(b2w)
plot(b2w)
@

@@

Below is the fit of a lognormal distribution to the same censored data set.
<<fitdistcensd2ln,fig=TRUE>>=
f2l<-fitdistcens(d2,"lnorm")
summary(f2l)
plot(f2l,leftNA=0)
@

Below is the fit of an exponential distribution.
<<fitdistcensd2e,fig=TRUE>>=
f2e<-fitdistcens(d2,"exp")
summary(f2e)
plot(f2e,leftNA=0)
@

@@

\section{Changing the optimization algorithm used to maximize the likelihood}

Sometimes the default algorithm used to maximize the likelihood fails to converge. It may then
be interesting to change some options of the function \texttt{optim} or to use another optimization
function than \texttt{optim} to maximize the likelihood.

\subsection{Changing the arguments passed to \texttt{optim}}

The argument optim.method may be used in the call to \texttt{fitdist} or
\texttt{fitdistcens}. It will internally be passed to \texttt{mledist} and to \texttt{optim}.
This argument may be fixed to \texttt{"Nelder-Mead"} (the robust Nelder and Mead method), 
\texttt{"BFGS"} (the BFGS quasi-Newton method), \texttt{"CG"} (a conjugate gradients method), 
\texttt{"SANN"} (a variant of simulated annealing) or \texttt{"L-BFGS-B"} (a modification of the BFGS 
quasi-Newton method which enables box constraints optimization). For the use of the last method the 
arguments \texttt{lower} and/or \texttt{upper} also have to be passed. More details on these optimization
functions may be found in the help page of \texttt{optim} from the package \texttt{stats}.

Below are examples of fits of a gamma distribution to non censored data with various options
of \texttt{optim}.

<<>>=
fitdist(x1,"gamma",optim.method="Nelder-Mead")
fitdist(x1,"gamma",optim.method="BFGS") 
fitdist(x1,"gamma",optim.method="L-BFGS-B",lower=c(0,0))
fitdist(x1,"gamma",optim.method="SANN")
@

\subsection{Supplying another optimization function}

You may also want to use another function than \texttt{optim} to maximize the likelihood.
This optimization function has to be specified by the argument \texttt{custom.optim} 
in the call to \texttt{fitdist} or
\texttt{fitdistcens}. But before that, it is necessary to customize this optimization function :
\texttt{custom.optim} function must have (at least) the following arguments,
\texttt{fn} for the function to be optimized, \texttt{par} for the initialized parameters. 
It is assumed that \texttt{custom.optim} should carry out a MINIMIZATION. Finally, it should return
at least the following components: \texttt{par} for the estimate, \texttt{convergence} for the convergence
code, \texttt{value} for \texttt{fn(par)} and \texttt{hessian}.

Below is an example of code written to customize \texttt{genoud} function from \texttt{rgenoud} package.
\begin{verbatim}
mygenoud <- function(fn, par, ...) 
{
   require(rgenoud)
   res <- genoud(fn, starting.values=par, ...)        
   standardres <- c(res, convergence=0)
   return(standardres)
}
\end{verbatim}

The customized optimization function may then be passed as the argument \texttt{custom.optim} 
in the call to \texttt{fitdist}
or \texttt{fitdistcens}. The following code may for example be used to fit a gamma distribution to the
non censored data \texttt{x1}. Note that in this example various arguments are also passed from \texttt{fitdist} 
to \texttt{genoud} :
\texttt{nvars}, \texttt{Domains}, \texttt{boundary.enforcement}, 
\texttt{print.level} and \texttt{hessian}.

\begin{verbatim}
fitdist(x1, "gamma", custom.optim=mygenoud, nvars=2,    
   Domains=cbind(c(0,0), c(10, 10)), boundary.enforcement=1, 
   print.level=1, hessian=TRUE)
\end{verbatim}
